---
title: "INFO 201 PS5: Exploring Carbon Emissions"
author: "Dan Sutton"
output: html_document
date: "2023-02-15"
---

```{r setup, include=FALSE} 
## Establishes code chunk behavior norms, loads tidyverse library, and loads the
##   data set into a variable.
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse) ## install.packages("tidyverse")
gm <- read_delim("data/gapminder.csv.bz2")
```

## Load and Check Data

The Gapminder data has `r nrow(gm)` rows and `r ncol(gm)` columns. The sample
table below suggests the dataset as a whole is in reasonably usable condition.
There are many missing pieces of data, which will be important to keep in mind
for the rest of the problem set.

```{r}
## Prints a sample of 5 rows.
gm %>% 
  sample_n(5)
```

## Descriptive Statistics

```{r}
## Creates the 'countries' dataframe, which contains each unique combination of
##   name, iso2, and iso3 variables.
countries <- gm %>% 
  select(name, iso2, iso3) %>% 
  unique()
```

Country Names: `r n_distinct(countries$name)`

ISO 2 Codes: `r n_distinct(countries$iso2)`

ISO 3 Codes: `r n_distinct(countries$iso3)`

```{r}
## Creates a dataframe that records how many names there are for each iso2.
namesPerIso2 <- countries %>% 
  select(name, iso2) %>% 
  group_by(iso2) %>% 
  summarize(totalNames = n_distinct(name)) %>% 
  arrange(desc(totalNames))
```

There is exactly one name for every ISO 2 code, with one exception. Namibia's
ISO 2 code appears to be "NA", which is also what denotes a missing ISO 2 code.

```{r}
## Shows that the iso2 value "NA" is paired with both Namibia and "NA" in the
##   name column.
countries %>% 
  select(name, iso2) %>% 
  unique() %>% 
  filter(is.na(iso2)) %>% 
  knitr::kable(col.names = c("Country Name", "ISO 2 Code"),
               align = c("l", "c"))

## Creates a dataframe that records how many iso3 codes there are for each name.
iso3PerName <- countries %>% 
  select(name, iso3) %>% 
  group_by(name) %>% 
  summarize(totalISO3 = n_distinct(iso3)) %>% 
  arrange(desc(totalISO3))

## Creates a dataframe showing the iso3 codes associated with "NA" name values.
iso3NA <- countries %>% 
  select(name, iso3) %>% 
  filter(is.na(name)) %>% 
  unique()
```

There are four different ISO 3 codes that correspond to the "NA" (missing value)
in the name column. Those codes are `r iso3NA$iso3`.

The earliest year in the data is `r min(gm$time, na.rm=TRUE)`, and the most
recent year is `r max(gm$time, na.rm=TRUE)`.

## CO2 Emissions

The following table shows how many missing values there are for CO2 and CO2 per
capita for each year. The last year that has any data for CO2 or CO2 per capita
is 2016; 2017-2019 contain values for some of the other columns, but have
missing values for CO2 and CO2 per capita for every country in the database.
Aside from that, the year with the most missing data for both CO2 and CO2 per
capita is 1960, and the amount of missing data goes down each year from there.

```{r}
## Creates a table as described in the above paragraph.
gm %>% 
  group_by(time) %>% 
  summarize(missing_co2 = sum(is.na(co2)),
            missing_co2_PC = sum(is.na(co2_PC))) %>% 
  knitr::kable(col.names = c("Year",
                             "Countries Missing CO2 Values",
                             "Countries Missing CO2 Per Capita Values"),
               align = c("l", "c", "c"))
```

The following plot shows the CO2 emissions over time for China, the United
States, Russia, India, Japan, and Germany.

A few interesting things to note:

* China's emissions have been rising dramatically for several decades, but
taper off towards the end of the available data.

* The United Stats has consistently had much higher emissions than the rest of
the countries in the set, generally rising from 1960 up until around 2010, when
carbon emissions start to drop.

* Russia's emissions show a sharp decline in the early 1990s, presumably from
the dissolution of the USSR splitting those emissions across several other
countries.

* The rest of the countries on the chart have lines that are
comparatively flat and low. India's emissions are growing, breaking them away
from this group, but their emissions are still half of that of the United
States, and a quarter of China's.

```{r}
## Creates a line plot of CO2 emissions over time for the countries specified
##   in the "countryPlot" vector.
countryPlot <- c("CHN", "USA", "RUS", "IND", "JPN", "DEU")

co2Emissions <- gm %>% 
  select(iso3, name, time, co2) %>% 
  filter(iso3 %in% countryPlot) %>%
  filter(time >= 1960 & time <= 2016) %>% 
  mutate(co2 = co2 / 1000000) %>% 
  filter(!is.na(co2))

ggplot(data = co2Emissions, aes(time, co2, color = iso3)) +
  geom_line() +
  labs(x = "Year",
       y = "CO2 emissions (millions of kilotons)",
       title = "CO2 emissions per year by country",
       color = "Country")
```

This line plot shows the same countries, this time measuring CO2 per capita
between 1960 and 2016.

* This chart reveals that even though China's absolute quantity of emissions per
year are twice that of the United States, that relationship is inverted when
considering the per capita measurement. This suggests that improvements that can
be made to emissions will be more efficient when applied in China as compared to
the United States.

* We see the same "drop" in emissions in Russia as we did in the previous table.

* India's emissions per capita are rising, but very slowly by comparison to
all the other countries on the list. India's rising population is likely just as
responsible for the recent increase in emissions as the increase per capita is.

```{r}
## Creates a line plot of CO2 emissions per capita over time for the countries
##   specified in the "countryPlot" vector.
co2Emissions_PC <- gm %>% 
  select(iso3, name, time, co2_PC) %>% 
  filter(iso3 %in% countryPlot) %>% 
  filter(time >= 1960 & time <= 2016) %>% 
  filter(!is.na(co2_PC))

ggplot(data = co2Emissions_PC, aes(time, co2_PC, color = iso3)) +
  geom_line() +
  labs(x = "Year",
       y = "CO2 emissions (metric tons per capita)",
       title = "CO2 emissions per capita per year by country",
       color = "Country")
```

```{r}
## Computes average CO2 emissions per capita, each year, for each region.
regionalCO2PC <- gm %>% 
  group_by(region, time) %>% 
  summarize(co2_PC_avg = mean(co2_PC, na.rm = TRUE))
```

Below is a plot showing the average CO2 emissions for each region in 1960 and
2016. A look at the complete data for the time between those years reveals what
you might expect from looking at just the plot below. The Americas and Europe
have been more responsible for emissions per capita than the rest of the world,
though the Americas shows what I expect is a misleading decline that prompts
further inquiry. Asia and Oceania have both seen a dramatic increase in
emissions per capita, and while Africa has also seen an increase in emissions,
the amount per capita is still far lower than any other region.

```{r}
## Creates a plot showing the regional CO2 emissions per capita for 1960 and
##   2016.
regionalCO2PCPlot <- regionalCO2PC %>% 
  filter(time == 1960 | time == 2016, !is.na(region))

ggplot(data = regionalCO2PCPlot, aes(region, co2_PC_avg, fill=factor(time))) +
  geom_col(position = "dodge2") +
  labs(x = "Region",
       y = "Average CO2 emissions (metric tons per capita)",
       title = "Average CO2 Emissions for each region in 1960 and 2016",
       fill = "Year")
```

The three highest and three lowest emitters, per capita, for each region in 2016
are found, combined, in the table below.

```{r}
## Creates tables showing the highest and lowest emitters per capita for each
##   region.
highEmitters <- gm %>% 
  group_by(region) %>% 
  filter(time == 2016, !is.na(region)) %>% 
  arrange(desc(co2_PC)) %>% 
  slice(1:3)

lowEmitters <- gm %>% 
  group_by(region) %>% 
  filter(time == 2016, !is.na(region)) %>% 
  arrange(co2_PC) %>% 
  slice(1:3)

co2PCLeaders <- c(unique(highEmitters$iso3), unique(lowEmitters$iso3))

gm %>% 
  filter(time == 2016, iso3 %in% co2PCLeaders) %>% 
  select(name, region, co2, co2_PC) %>% 
  arrange(region, desc(co2_PC)) %>% 
  knitr::kable(col.names = c("Country",
                             "Region",
                             "CO2 Emissions",
                             "CO2 Emissions Per Capita"),
               align = c("l", "l", "r", "r"))
```

## GDP Per Capita

In the graph below, we can see that GDP per capita is a poor predictor of life
expectancy. Below about $5,000 GDP per capita, it is difficult to predict
whether life expectancy is low or high, though there is a vague positive
correlation. Above $5,000, there is a very slight positive trend in life
expectancy with increasing GDP per capita, but almost every country is near or
above 70 years, with a small number of exceptions.

Europe dominates with respect to both overall life expectancy and GDP per
capita. The Americas generally have higher life expectancies than the rest of
the world, thought not necessarily much higher GDP per capita.

```{r}
## Creates a scatterplot for life expectancy vs GDP per capita in 1960.
GDPvsLE1960 <- gm %>% 
  filter(time == 1960, !is.na(lifeExpectancy), !is.na(GDP_PC)) %>% 
  mutate(totalPopMil = totalPopulation / 1000000,
         GDP_PC_k = GDP_PC / 1000)

ggplot(data = GDPvsLE1960,
       aes(GDP_PC_k, lifeExpectancy, color=region, size=totalPopMil)) +
  geom_point() +
  labs(x = "GDP per capita (thousands of dollars)",
       y = "Life expectancy (years)",
       title = "Life expectancy vs GDP per capita, 1960",
       color = "Region",
       size = "Population, in millions")
```

While the specific numbers have changed, the trend largely remains the same. In
2019, Asian countries seem to have improved the most across the board, at least
with respect to life expectancy. The highest life expectancy values are higher
than before, as are GDP per capita values.

Of note is that in 2019, there is data for a single country, ISO 3 code KOS,
that is not allocated to any of the 5 regions defined in the data. This is the
ISO 3 code for Kosovo, a region in southeast Europe that is only recognized by
part of the international community. I would expect Kosovo to be part of the
Europe region, but perhaps this disputed status as a country has something to
do with its "regionless" status.

```{r}
## Creates a scatterplot for life expectancy vs GDP per capita in 2019.
GDPvsLE2019 <- gm %>% 
  filter(time == 2019, !is.na(lifeExpectancy), !is.na(GDP_PC)) %>% 
  mutate(totalPopMil = totalPopulation / 1000000,
         GDP_PC_k = GDP_PC / 1000)

ggplot(data = GDPvsLE2019,
       aes(GDP_PC_k, lifeExpectancy, color=region, size=totalPopMil)) +
  geom_point() +
  labs(x = "GDP per capita (thousands of dollars)",
       y = "Life expectancy (years)",
       title = "Life expectancy vs GDP per capita, 2019",
       color = "Region",
       size = "Population, in millions")
```

Average life expectancy has increased for all regions between 1960 and 2019.
Asia and Africa have both seen the most improvement, while the Americas and
Oceania still see significant, though somewhat less pronounced, improvements.
Europe (and Kosovo) have seen the smallest increase. All of this correlates well
with the scatterplot.

```{r}
## Computes the average life expectancy for each region over time.
LE <- gm %>% 
  group_by(region, time) %>% 
  summarize(avg_LE = mean(lifeExpectancy, na.rm=TRUE)) %>% 
  mutate(LE_growth = avg_LE - lag(avg_LE))

LE_1960 <- LE %>% 
  filter(time == 1960) %>% 
  mutate(avg_LE_1960 = avg_LE) %>% 
  select(region, avg_LE_1960)

LE_2019 <- LE %>% 
  filter(time == 2019) %>% 
  mutate(avg_LE_2019 = avg_LE) %>% 
  select(region, avg_LE_2019)

LE_Joined <- full_join(LE_1960, LE_2019)

LE_Joined %>% 
  mutate(LE_Change = avg_LE_2019 - avg_LE_1960) %>% 
  arrange(desc(LE_Change)) %>% 
  knitr::kable(col.names = c("Region",
                             "Average Life Expectancy, 1960",
                             "Average Life Expectancy, 2019",
                             "Change in Life Expectancy"),
               align = c("l", "c", "c", "c"))
```

The table below shows the one-year improvements in growth, ordered from largest
growth to smallest growth. Region "NA" (Kosovo) has been excluded from this
already otherwise massive table.

In general, the largest positive changes in life expectancy happen mostly in
places where the current average life expectancy is in the 50s and low 60s.
This suggests that the factors that cause average life expectancy to increase
above 50 are largely the same factors that will eventually push it up to 70 or
so.

The smallest growth (in some cases, negative growth) mostly seem to happen near
or above 70, and near 50. That suggests that 50 and 70 are stable average life
expectancy values, and that it is challenging to push values higher than 70 or
lower than 50.

```{r}
## Creates a table that ranks the life expectancy growth for each region each
##   year from highest growth to lowest growth.
LE %>% 
  arrange(desc(LE_growth)) %>% 
  filter(!is.na(LE_growth), !is.na(region)) %>% 
  knitr::kable(col.names = c("Region",
                             "Year",
                             "Average Life Expectancy",
                             "Change in Life Expectancy"),
               align = c("l", "c", "c", "c"))
```

```{r}
## Creates a histogram of GDP per capita values for 1960 and 2019.
GDP_PC_hist <- gm %>% 
  filter(time == 1960 | time == 2019, !is.na(GDP_PC)) %>% 
  group_by(time)

ggplot(data = GDP_PC_hist, aes(GDP_PC, fill = factor(time))) +
  geom_histogram(position="dodge") +
  labs(x = "GDP per capita",
       y = "Number of countries",
       title = "Histogram of GDP per capita values, 1960 and 2019",
       fill = "Year")
```

```{r}
## Creates a table to show how the United States fares on life expectancy as
##   compared to the rest of the world.
gm %>%
  filter(!is.na(name), !is.na(lifeExpectancy)) %>% 
  group_by(time) %>% 
  mutate(LE_rank = rank(desc(lifeExpectancy)),
         total_countries = n(),
         LE_rank_percent = LE_rank / total_countries * 100) %>% 
  ungroup() %>% 
  filter(time == 1960 | time == 2019, iso3 == "USA") %>% 
  select(name, time, lifeExpectancy, LE_rank, total_countries,
         LE_rank_percent) %>% 
  knitr::kable(col.names = c("Country",
                             "Year",
                             "Life Expectancy",
                             "Life Expectancy Rank",
                             "Number of Countries",
                             "Percentile"))
```

I spent *very roughly* ten hours working on this assignment, but I did not keep
close track, and some of the time I spent working on the assignment were going
down ggplot aesthetic rabbit holes, much of which was learning about techniques
that weren't applicable in this case.